---
title: "Object oriented programming in R"
output: 
  html_document: 
    toc: yes
---

```{r, echo = FALSE}
library(sloop)
```

## Objects in R

We first need to talk about type of objects. Everthing in R are objects, but there are different types of objects.

- Base types, such as integers, characters etc
- S3, e.g., `data.frame`, `tibble`
- S4, e.g., the matrices from `Matrix` package (getting less popular nowadays)
- RC, e.g., not popular
- R6, e.g., the state of the art OOP system


We will use the `sloop` library to help us understand some internal R structures.


### Base types

`otype` shows the type of an object.

```{r}
otype(1:10)
otype("a")
```

```{r}
x <- matrix(1:2, nrow = 2, ncol = 2)
otype(x)
```


### S3
```{r}
otype(mtcars)
class(mtcars)
```

```{r}
fit <- lm(mpg ~ wt, data = mtcars)
otype(fit)
class(fit)
```

### S4
```{r}
M <- Matrix::Matrix(rnorm(12), 3, 4)
otype(M)
```

```{r}
M@Dim # `@` is the signature of S4 objects
```

```{r}
class(M)
```


### R6

```{r}
# it is a package that we will learn later
library(callr)
```

```{r}
bg_task <- r_bg(function() Sys.sleep(60))
otype(bg_task)
```

```{r}
bg_task$is_alive() # `$` to call methods
```

## OOP

Orject oriented programming is a difficult topic. We will breifly discuss two concepts in OOP, i.e., polymorphism (S3/S4) and encapsulation (R6).


### Polymorphism

So what is it? Literally, it means many forms

```{r}
x <- rnorm(10)
y <- rnorm(10)
fit <- lm(y ~ x)
```

```{r}
summary(x) # summary.default(x)
```

```{r}
summary(fit) # summary.lm(fit)
```

In above, the function `summary` takes different forms and dispatches different results based on the class of the input.

```{r}
otype(x)
```

```{r}
otype(fit)
class(fit)
```

`summary` is a generic function which dispatches other functions based on the class of the object.
```{r}
# list all the method dispatches of summary
s3_methods_generic("summary")
```

Another example of dispatching
```{r}
ta <- table(mtcars$cyl, mtcars$am)
otype(ta)
summary(ta)  # summary.table(ta)
```


#### Extend existing generic

Check if a function is generic
```{r}
# `length` is a primitive generic function
ftype(length)
```
```{r}
# `mean` is a S3 generic function
ftype(mean)
```
```{r}
# lm is not generic
ftype(lm)
fit <- lm(mpg ~ wt , data = mtcars)
```


As `length` is a primitive generic function, it dispatches different versions of `length` for different inputs.

```{r}
s3_methods_generic("length")
```


`length` returns the length of a vector, not the string length
```{r, error = TRUE}
s <- "it is long string"
otype(s)
length(s)
```

The method should extend the signatures of the generic.
```{r}
# To define a `length` method for `string`.
length.string <- function(x) {
  nchar(x)
}
```

```{r}
s <- structure("it is long string", class = "string")
# it is the same as
# s <- "it is long string"
# class(s) <- "string"
otype(s)
class(s)
length(s) # length.string(s)
```

Another example, "print" function for string

```{r}
ftype(print)
```
```{r}
print.string <- function(x, ..., max_length = 40) {
  cat(stringr::str_trunc(x, max_length))
  cat("\n")
}
```

```{r}
s <- structure("it is long string jflkdj aflkjdklafj kldasjf fdasjklfjdlkas fdajflkdas", class = "string")
s
print(s, max_length = 60)
```


##### Difference between primitive and S3 generic function.

- primitive generic cannot be extended to base types.

```{r}
otype(LETTERS)
ftype(sum)
```
```{r, error = TRUE}
sum(LETTERS)
```

```{r, echo = FALSE}
# it is not possible, we cannot extend `sum` for `character` objects 
# because `sum` is primitive and `character` is base class
sum.character <- function(..) {
  
}
```

- On the other hand, S3 could be extended to base types. 

```{r}
ftype(mean)
```

```{r}
mean.character <- function(x, na.rm = FALSE, ...) {
  # please ignore the implmentation detail here
  # it is just a way to define mean for letters
  intToUtf8(round(mean(sapply(x, utf8ToInt), na.rm = na.rm)))
}
mean(c("A", "G", "H"))
```

```{r}
# without the extension
rm(mean.character)
mean(c("A", "G", "H"))
```

#### Create your own generic

```{r}
center <- function(x) {
  UseMethod("center")
}
```

```{r}
ftype(center)
```

Then we define two versions of center based on the class.
```{r}
center.skewed <- function(x) {
  mean(x, trim = 0.2)
}

center.symmetric <- function(x) {
  mean(x)
}
```

Fallback, if nothing works
```{r}
center.default <- function(x) {
  median(x)
}
```


```{r}
a <- rnorm(10)
class(a) <- "skewed"
center(a) # 80% trim mean:  center.skewed(a)
mean(a, trim = 0.2)
```

```{r}
class(a) <- "symmetric"
center(a) # ordinary mean:  center.symmetric(a)
mean(a)
```

```{r}
class(a) <- "anything_else"
center(a) # median
median(a)
```

```{r}
b <- rnorm(10)
center(b)  # center.default(b); median(b)
```


### Encapsulation

R6 has two special properties:

- It uses the encapsulated OOP paradigm, which means that methods belong to objects, not generics, and you call them like `object$method()`.
- R6 objects are mutable, which means that they are modified in place, and hence have reference semantics.


`R6Class` is not defined in base R, it is located in the package `R6`. There only one function
that we would ever use `R6::R6Class`.

A R6 class provides a general structure of an R6 object.
The following example shows the two most important arguments to `R6Class()`

- The first argument is the classname. It’s not strictly needed, 
- The second argument, `public`, supplies a list of methods (functions) and fields (anything else) that make up the public interface of the object. Methods can access the methods and fields of the current object via `self$`.

```{r}
Accumulator <- R6::R6Class(
  "Accumulator",
  public = list(
    sum = 0,
    add = function(x = 1) {
      self$sum <- self$sum + x
      invisible(self)  # chain method calls
    }
  )
)
```

You construct a new object from the class by calling the `new()` method. 


```{r}
x <- Accumulator$new()
```

You can then call methods and access fields with `$`:
```{r}
x$add(4)
```

```{r}
x$sum
```


#### Method chaining

Side-effect R6 methods should always return self invisibly. This returns the "current" object and makes it possible to chain together multiple method calls:

```{r}
x$add(10)$add(10)$sum
```

For readability
```{r}
x$
  add(10)$
  add(10)$
  sum
```

#### Initialization

`$initialize()` overrides the default behaviour of `$new()`. 
For example, the following code defines an Person class with fields `$name` and `$age`.

```{r}
Person <- R6::R6Class(
  "Person", 
  public = list(
    name = NULL,
    age = NA,
    initialize = function(name, age = NA) {
      stopifnot(is.character(name), length(name) == 1)
      stopifnot(is.numeric(age), length(age) == 1)
      self$name <- name
      self$age <- age
    }
  )
)
```

```{r}
randy <- Person$new("Randy", 33)
randy
```

```{r, error = TRUE}
randy <- Person$new("Randy", "thirth-three")
```

#### Printing

The `$print` method allows you to override the default printing behaviour.

```{r}
Person <- R6::R6Class(
  "Person", 
  public = list(
    name = NULL,
    age = NA,
    initialize = function(name, age = NA) {
      self$name <- name
      self$age <- age
    },
    print = function(...) {
      cat("Person: \n")
      cat("  Name: ", self$name, "\n", sep = "")
      cat("  Age:  ", self$age, "\n", sep = "")
      invisible(self)
    }
  )
)
```

```{r}
randy <- Person$new("Randy", 33)
randy
```

#### Mutability

Unlikes most R objects, R6 objects are mutable.

```{r}
# list is immutable
randy <- list(name = "Randy", age = 33)
randy_second <- randy
randy_second$age <- 18
randy$age
```

```{r}
# R6 object is mutable
randy <- Person$new("Randy", 33)
randy_second <- randy
randy_second$age <- 18
randy$age
```

To make a clone, we need to use the `$clone` method
```{r}
randy <- Person$new("Randy", 33)
randy_second <- randy$clone()
randy_second$age <- 18
randy$age
```
PS: `$clone()` does not recursively clone nested R6 objects. If you want that, you’ll need to use `$clone(deep = TRUE)`

#### Mutable default argument

A similar issue could be raised in Python.

```{python}
class Foo():
  x = []
  def add(self, y):
    self.x.append(y)
```

```{python}
a = Foo()
b = Foo()
a.x
b.x
```


```{python}
a.add(1)
a.add(2)
b.x
```

We need to initlize the default variables in `__init__`
```{python}
class Foo():
  x = None
  def __init__(self):
    self.x = []
  def add(self, y):
    self.x.append(y)
```

To demonstrate the same issue
```{r}
Store <- R6::R6Class(
  "Store", 
  public = list(
    owner = Person$new("Randy"),
    transfer = function(name) {
      self$owner$name <- name
      invisible(self)
    }
  )
)
```

```{r}
bookstore <- Store$new()
cafe <- Store$new()
```

```{r}
bookstore$transfer("John")
```

Since both `bookstore` and `cafe` have the same object `Person$new("Randy")`, the owner name of `cafe` is also changed.

```{r}
cafe$owner
```

```{r}
Store <- R6::R6Class(
  "Store", 
  public = list(
    owner = NULL,
    initialize = function() {
      self$owner <- Person$new("Randy")
    },
    transfer = function(name) {
      self$owner$name <- name
      invisible(self)
    }
  )
)
```


```{r}
bookstore <- Store$new()
cafe <- Store$new()
```

```{r}
bookstore$transfer("John")
```

```{r}
cafe$owner
```


#### Finalizer

Since R6 is mutable in nature, it is useful to have a `$finalizer()` method called after all the references to an object are destroyed.

See https://adv-r.hadley.nz/r6.html#finalizer


# Reference

Advanced R: https://adv-r.hadley.nz/oo.html
