---
title: "Object oriented programming in R"
output: 
  html_document: 
    toc: yes
---

```{r, echo = FALSE}
library(sloop)
```

## Objects in R

We first need to talk about type of objects. Everthing in R are objects, but there are different types of objects.

- Base types, such as integers, characters etc
- S3, e.g., `data.frame`, `tibble`
- S4, e.g., the matrices from `Matrix` package (getting less popular nowadays)
- RC, e.g., not popular
- R6, e.g., the state of the art OOP system


We will use the `sloop` library to help us understand some internal R structures.


### Base types

```{r}
otype(1:10)
otype("a")
otype(mean)
```


### S3
```{r}
otype(mtcars)
s3_class(mtcars)
```

```{r}
fit <- lm(mpg ~ wt, data = mtcars)
otype(fit)
s3_class(fit)
```


Objects of other types could also have S3 classes

```{r}
x <- matrix(1:2, nrow = 2, ncol = 2)
otype(x)
s3_class(x)
```

### S4
```{r}
library(Matrix)
M <- Matrix(rnorm(12), 3, 4)
otype(M)
```

```{r}
M@Dim  # `@` is the signature of S4 objects
```

### R6

```{r}
# it is a package that we will learn later
library(callr)
```

```{r}
bg_task <- r_bg(function() Sys.sleep(60))
otype(bg_task)
```

```{r}
bg_task$is_alive()  # `$` to call methods
```

## OOP

Orject oriented programming is a difficult topic. We will breifly discuss two concepts in OOP, i.e., polymorphism (S3) and encapsulation (S4, R6).


### Polymorphism

So what is it? Literally, it means many forms

```{r}
x <- rnorm(10)
y <- rnorm(10)
fit <- lm(y ~ x)
```

```{r}
summary(x)   # summary.default(x)
```

```{r}
summary(fit)  # summary.lm(fit)
```

In above, the function `summary` takes different forms and dispatches different results based on the class of the input.

```{r}
otype(x)
s3_class(x)
```


```{r}
otype(fit)
s3_class(fit)
```




`summary` is a generic function which dispatches other functions based on the class of the object.
```{r}
# list all the method dispatches of summary
s3_methods_generic("summary")
```

#### Reuse exsiting generic


`sum` is a generic function

```{r}
sum
```

We cannot sum up letters.
```{r, error = TRUE}
x <- c("a", "b", "c")
sum(x)
```

```{r}
class(x) <- "myclass"
s3_class(x)
```

```{r}
sum.myclass <- function(..., na.rm = FALSE) {
  paste0(..., collapse = "")
}
```

```{r}
sum(x)
```


#### Create your own generic

```{r}
center <- function(x) {
  UseMethod("center")
}
```

Then we define two versions of center based on the class.
```{r}
center.foo <- function(x) {
  mean(x, trim = 0.2)
}

center.bar <- function(x) {
  mean(x)
}
```

Fallback, if nothing works
```{r}
center.default <- function(x) {
  median(x)
}
```



```{r}
a <- rnorm(10)
class(a) <- "foo"
center(a)  # 80% trim mean
mean(a, trim = 0.2)
```

```{r}
class(a) <- "bar"
center(a)  # ordinary mean
mean(a)
```

```{r}
class(a) <- "anything"
center(a)  # ordinary mean
median(a)
```


# Reference

Advanced R: https://adv-r.hadley.nz/oo.html
