---
title: "Python interoperability"
output: 
  html_document: 
    toc: yes
---


# R vs Python

Why not both?


## Setup

### Step 1
To use python in R, we need to install the R package `reticulate`.
```{r}
# make sure it is 1.14 or above
library(reticulate)
```

### Step 2


Display python environment detected. You will be prompted to install `miniconda` if no suitable python is detected.

```{r, eval = FALSE}
py_config()
```

In this class, we are going to focus on `miniconda.` If your system has another workable python environment. You will need to manually install `miniconda` with
```{r, eval = FALSE}
reticulate::install_miniconda()
```
and specify to use `miniconda` after loading `reticulate`, like this
```{r, eval = FALSE}
library(reticulate)
use_miniconda()  # only needed if your computer has multiple installations of python.
```

Visit [reticulate](https://rstudio.github.io/reticulate/articles/versions.html) website if you want to configure R to use different version of python then miniconda.

## Ways to use python

There are a variety of ways to integrate Python code into your R projects:

- Python in R Markdown 

```{python}
import random
x = random.random()
```

```{r}
# to access Python objects in R
py$x
```


- Importing Python module

The `import()` function enables you to import any Python module and call its functions directly from R. 

```{r}
random <- import("random")
y <- random$random()
```

```{python}
# to access R object in python
r.y
```

The `import_builtins()` function enables to access the [built in functions](https://docs.python.org/3/library/functions.html#built-in-funcs).

```{r}
builtins <- import_builtins()
a <- builtins$range(5L)

builtins <- import_builtins()
builtins$len(a)
```


- Sourcing Python scripts: The `source_python()` function enables you to source a Python script the same way you would source() an R script (Unless `envir = NULL`, Python functions and objects defined within the script become directly available to the R session).


```{r}
source_python("script.py", envir = NULL)
py$z
```

```{r}
# I personally don't recommand it
source_python("script.py")
z
```


- Python REPL: `repl_python()`


## Python basics

https://learnxinyminutes.com/docs/python/
https://www.pyscoop.com/learn-python-in-10-minutes/


## Type conversions

https://rstudio.github.io/reticulate/#type-conversions

By default when Python objects are returned to R they are converted to their equivalent R types.

```{r}
random <- import("random")
(x <- random$random())
class(x)
```

However, if you'd rather make conversion from Python to R explicit and deal in native Python objects by default you can pass `convert = FALSE` to the import function

```{r}
random <- import("random", convert = FALSE)
(x <- random$random())
class(x)
```

We cannot work with native Python objects directly.
```{r, error = TRUE}
x + 1
```

The function `py_to_r` converts native Python objects to R objects.
```{r}
# convert x to R object first
py_to_r(x) + 1
```

Let's check another numpy example.
```{r}
np <- import("numpy", convert = FALSE)

# do some array manipulations with NumPy
a <- np$array(c(1:4))
sum <- a$cumsum()

# convert to R explicitly at the end
py_to_r(sum)
```

R Objects are converted to Python objects when they are passed to Python functions.

```{python}
import sys
def abssum(x):
    print("received a {} object\n".format(type(x)))
    return sum((abs(z) for z in x))
```

```{r}
x <- rnorm(10)
# x is implicitly to a native Python list
py$abssum(x)
```

