---
title: "Functional programming in R"
output: 
  html_document: 
    toc: yes
---

You'll learn some of the powerful programming tools provided by `purrr`, one of the tidyverse core packages.

```{r, message = FALSE}
library(tidyverse)
```


## Pipe operator

We are going to detour a bit to talk about pipe operator.
Pipe operator is a powerful way to make your code much cleaner and readible by

- structuring sequences of data operations left-to-right (as opposed to from the inside and out),
- avoiding nested function calls,
- minimizing the need for local variables and function definitions, and
- making it easy to add steps anywhere in the sequence of operations.

The operators pipe their left-hand side values forward into expressions that
appear on the right-hand side, i.e. one can replace `f(x)` with 
`x %>% f()`, where `%>%` is the (main) pipe-operator. When coupling 
several function calls with the pipe-operator, the benefit will become
more apparent. Consider this pseudo example:

```{r, eval = FALSE}
raw_data <- read.csv("/path/to/data/file.csv")
sub_data <- subset(raw_data, variable_a > x)
new_data <- transform(sub_data, variable_c = variable_a / variable_b)
the_data <- head(new_data, 100)
```
However, the local variables are really not necessary, so one liner could be

```{r, eval = FALSE}
the_data <- head(
  transform(
    subset(
      read.csv("/path/to/data/file.csv"), 
    variable_a > x), 
  variable_c = variable_a / variable_b), 
100)
```

But that is too hard to read and reason. If we use pipe operator,

```{r, eval = FALSE}
the_data <- read.csv("/path/to/data/file.csv") %>%
  subset(variable_a > x) %>%
  transform(variable_c = variable_a / variable_b) %>%
  head(100)
```

Hint: In RStudio, you could use `Ctrl + Shift + M` (or `Cmd + Shift + M`) to insert the pipe operator.


### Basic usage

* `x %>% f` or `x %>% f()` is equivalent to `f(x)`
* `x %>% f(y)` is equivalent to `f(x, y)`
* `x %>% f %>% g %>% h` is equivalent to `h(g(f(x)))`


```{r}
choose(5, 3)
5 %>% choose(3)
```


### The argument placeholder

* `x %>% f(y, .)` is equivalent to `f(y, x)`
* `x %>% f(y, z = .)` is equivalent to `f(y, z = x)`

```{r}
3 %>% choose(5, .)
3 %>% choose(5, k = .)
```

### Re-using the placeholder

It is straightforward to use the placeholder several times
in a right-hand side expression. However, when the placeholder
only appears in a nested expressions magrittr will still apply
the first-argument rule. The reason is that in most cases this
results more clean code. 

`x %>% f(y = nrow(.), z = ncol(.))` is equivalent to 
   `f(x, y = nrow(x), z = ncol(x))`

```{r}
5 %>% choose(. - 2)
```

The behavior can be
overruled by enclosing the right-hand side in braces:

`x %>% {f(y = nrow(.), z = ncol(.))}` is equivalent to 
   `f(y = nrow(x), z = ncol(x))`

```{r}
list(n = 5, k = 3) %>% {choose(.$n, .$k)}
```


## What is functional programming?

> In computer science, functional programming is a programming paradigm—a style of building the structure and elements of computer programs—that treats computation as the evaluation of mathematical functions and avoids changing-state and mutable data.

To put is simple, it is a programming style without side effects.


```{python, eval = FALSE}
def f(x):
  x["a"] = 2
  return x

x = {"a": 1}
f(x) == x
```

```{r}
f <- function(x) {
  x$a <- 2
  x
}

x <- list(a = 1)
isTRUE(all.equal(f(x), x))
```


## Is R a functional language?

R, at its heart, is a functional programming language. Most base R functions are pure, with a few notable exceptions:

- `library()` which loads a package, and hence modifies the search path.

- `setwd()`, `Sys.setenv()`, `Sys.setlocale()` which change the working directory, environment variables, and the locale, respectively.

- `plot()` and friends which produce graphical output.

- `write()`, `write.csv()`, `saveRDS()`, etc. which save output to disk.

- `options()` and `par()` which modify global settings.

- `S4` related functions which modify global tables of classes and methods.

- Random number generators which produce different numbers each time you run them

Almost everything else in R are functions.
```{r}
x <- 1
y <- 2
x + y
`+`(x, y)

a <- 3:5
a[3]
`[`(a, 3)

z <- 1
`<-`(z, 1)
```

## R is also very dynamic

We could change the body of previously defined functions
```{r}
add <- function(x, y) x + y
add(3, 1)

body(add) <- quote(x - y)
add(3, 1)
```

We could also change the frame that a function was defined at.

```{r}
y <- 2
addy <- function(x) {
  x + y
}
addy(3)
```

```{r}
foo <- function(x) {
  y <- 3
  addy(x)
}

foo(3)
```
`foo(3)` returns 3 because `addy` was defined in the global environment. But we could do something like this

```{r}
foo <- function(x) {
  y <- 3
  environment(addy) <- environment()
  addy(x)
}

foo(3)
```
`environment(addy) <- environment()` changes lexical scope of the function `addy` from global to the calling frame of `foo`.


## Revisit for loops

Suppose we want to compute the sample medians for each column of the data frame

```{r}
df <- tibble(
  a = rnorm(10),
  b = rnorm(10),
  c = rnorm(10),
  d = rnorm(10)
)
# if you want to be fancy
df <- tibble(!!!rerun(4, rnorm(10)) %>% set_names(letters[seq_len(length(.))]))
```

A naive way to do so
```{r}
median(df$a)
median(df$b)
median(df$c)
median(df$d)
```

```{r}
# a base R for loop
output <- vector("double", ncol(df)) # 1. output
for (i in seq_along(df)) { # 2. sequence
  output[[i]] <- median(df[[i]]) # 3. body
}
output
```

### Base R solution

```{r}
lapply(df, median)
sapply(df, median)
```
These functions are great but they are not far less powerful then the next function `map`.


## Using `purrr`

THe `purrr` package (part of `tidyverse`) provides a function `map`.

The basic syntax for `map` is
```
map(ITERATABLE, A_FUNCTION)
```
or
```
ITERATABLE %>% map(A_FUNCTION)
```

* `ITERATABLE`'s are objects like vectors, list and data frame.
* `A_FUNCTION` is a function, an expression, or a list of element names


```{r}
df %>% map(median)
```

Consider the following naive function which computes factorials
```{r}
fac <- function(n) {
  if (n == 1) {
    return(1L)
  } else {
    return(fac(n - 1L) * n)
  }
}
fac(5)
```

Note that `fac` is not vectorized, it means that it doesn't handle vector input.
```{r, error = TRUE}
fac(1:5)
```

```{r}
1:5 %>% map(fac)
```
is almost equivalent to `c(fac(1), fac(2), fac(3), fac(4), fac(5))`


### Passing additional arguments

```r
mean(..., trim = 0.5)
```
computes trimed mean of the input.

```{r}
df %>% map(mean, trim = 0.5)
```

### Inline function / expressions

```{r}
df %>% map(function(x) x^2)
```
or
```{r}
df %>% map(function(.) .^2)
```
or simply

```{r}
df %>% map(~ .^2)
```

### Name and position shortcuts

The `map` function also provide some shortcuts to extract elements from lists or vectors.

```{r}
roster <- list(
  list(name = "Peter", id = 1231, grades = list(STA141A = "A", STA141B = "A")),
  list(name = "Mary",  id = 4213, grades = list(STA141A = "B", STA141B = "B"))
)
```

```{r}
roster %>% map("name")
```

## Multi-level extraction
```{r}
roster %>% map(list("grades", "STA141B"))
```


## Type-specific map

`map()` always returns a list, even if all the elements have the same flavor and are of length one. But in that case, you might prefer a simpler object.

- `map_lgl()` makes a logical vector.
- `map_int()` makes an integer vector.
- `map_dbl()` makes a double vector.
- `map_chr()` makes a character vector.

```{r}
df %>% map_dbl(mean)
```

```{r}
1:5 %>% map_int(fac)
```

```{r}
1:5 %>% map_lgl(~ . == 2L)
```


## Enumerating over the input

```{r}
eleven_to_twenty <- c(11, 12, 13)
b <- c(4, 5, 6)
eleven_to_twenty %>% imap_dbl(function(v, i) v + b[i])
# equivalently
eleven_to_twenty %>% imap_dbl(~ .x + b[.y])
```


## Filtering

```{r}
eleven_to_twenty <- 11:20
eleven_to_twenty %>% pluck(1) # equivalent to eleven_to_twenty[1]

```

```{r}
a_to_e <- list(lowers = letters[1:5], uppers = LETTERS[1:5])
a_to_e %>% pluck("lowers") # equivalent to a_to_e["lowers"]
a_to_e %>% pluck("lowers", 2) # equivalent to a_to_e["lowers"][2]
```

## Keep and drop elements

```{r}
is_even <- function(x) x %% 2 == 0
eleven_to_twenty %>% keep(is_even)
# or equivalently
eleven_to_twenty %>% keep(~ . %% 2 == 0)

a_to_e %>% keep(~ "eleven_to_twenty" %in% .)
```

```{r}
eleven_to_twenty %>% keep(~ !is_even(.))
eleven_to_twenty %>% keep(negate(is_even))
eleven_to_twenty %>% discard(is_even)
```

## Summerizing

```{r}
# is every element in `eleven_to_twenty` even?
eleven_to_twenty %>% every(is_even) # more efficient then all(is_even(eleven_to_twenty))
# are some elements in `eleven_to_twenty` even?
eleven_to_twenty %>% some(is_even) # more efficient then any(is_even(eleven_to_twenty))

eleven_to_twenty %>% has_element(12) # 12 is eleven_to_twenty double
eleven_to_twenty %>% has_element(12L)

# the first even number is
eleven_to_twenty %>% detect(is_even)
# the first even number divisible by 7 is
eleven_to_twenty %>% detect(~ . %% 7 == 0)

# get the index of the first element  divisible by 7, then extract it
eleven_to_twenty %>%
  detect_index(~ . %% 7 == 0) %>%
  pluck(eleven_to_twenty, .)
```


## Modify elements selectively

```{r}
# eleven_to_twenty %>% modify(~ . + 1) results in an error because eleven_to_twenty is an integer vector
eleven_to_twenty %>% modify(~ . + 1L)

eleven_to_twenty %>% modify_if(is_even, ~ -.)
eleven_to_twenty %>% modify_if(~ !is_even(.), ~ -.)

eleven_to_twenty %>% modify_at(c(1, 3, 5), ~ . + 2L)
```


## Reshape lists

```{r}
a_to_e %>% flatten_chr()
```

```{r}
a_to_e %>%
  transpose() %>%
  map(~ str_c(.$lowers, .$uppers))
```

## Add elements to lists

```{r}
li <- list(a = 1, b = 2)
li %>%
  append(list(c = 3, d = 3)) %>%
  prepend(list(x = 0, y = 0)) %>%
  splice(u = 3, v = 4)
```


## Map more than one argument

Two arguments
```{r}
a <- c(7, 8, 9)
b <- c(2, 3, 4)

map2(a, b, choose)
a %>% map2(b, choose)
b %>% map2(a, ., choose)
# or equivalently
map2(a, b, ~ choose(.x, .y))
map2(a, b, function(x, y) choose(x, y))
```

More than two arguments
```{r}
l <- list(
  a = c(1, 2, 3),
  b = c(5, 4, 8),
  c = c(4, 7, 9)
)

l %>% pmap_dbl(function(a, b, c) a / (b + c))
```
 
## Cross

The `map` function applies to the inputs elementwisely. Sometimes, we wish to consider all combinations of the inputs. The `cross*` functions are designed to handle those situations.

`cross2` usually is used together with `lift`
```{r}
a <- c(7, 8, 9)
b <- c(2, 3, 4)
choose(7, 3)
lift(choose)(list(7, 3))
cross2(a, b) %>% map_dbl(lift(choose))
```

```{r}
li <- list(
  a = c(7, 8, 9),
  b = c(2, 3, 4)
)
li %>%
  set_names(c("n", "k")) %>%
  cross() %>%
  map(lift(choose))
```

```{r}
dat <- tibble(
  x = seq(-1, 1, length = 10),
  y = seq(-1, 1, length = 10)
)


# we wish to plot a heat map of sqrt(x^2 + y^2)

dat %>% mutate(z = sqrt(x^2 + y^2)) # wrong

dat %>%
  cross_df() %>%
  mutate(z = sqrt(x^2 + y^2)) %>%
  ggplot(aes(x, y, fill = z)) +
  geom_tile()
```


## Reduce iteratable

The `reduce` function combines the values of a vector by iteratively applying a binary function.

```{r}
x <- 1:4
x %>% reduce(`+`)
# it means ((1 + 2) + 3) + 4
1:4 %>% reduce(`+`, .dir = "backward")
# it means 1 + (2 + (3 + 4))

4:1 %>% reduce(choose)
# it means choose(choose(choose(4, 3), 2), 1)
```

It is most useful if we wants to row bind or columm bind a list of data frames.

```{r}
dfs <- list(
  tibble(name = "Peter", age = 23),
  tibble(name = "Paul", age = 19),
  tibble(name = "Mary", age = 22)
)
dfs %>% reduce(bind_rows)
```



```{r}
# shortcircuit a reduction
sum_until <- function(x, y) {
  out <- x + y
  if (out > 30) {
    done(out)
  } else {
    out
  }
}

1:10 %>% reduce(sum_until)
```

`accumulate` works similarly but keep all the imediate calculations.

```{r}
1:10 %>% accumulate(`+`)
```

```{r}
1:10 %>% accumulate(sum_until)
```

# A example of linear regression

```{r}
mtcars %>%
  nest(data = -cyl)

# or more readible
mtcars %>%
  group_by(cyl) %>%
  nest() %>%
  mutate(model = map(data, ~ lm(mpg ~ wt, data = .))) %>%
  transmute(r_sq = model %>% map(summary) %>% map_dbl("r.squared"))
```


# References

- R for Data Science https://r4ds.had.co.nz/iteration.html
- purrr tutorial https://jennybc.github.io/purrr-tutorial/
- purrr cheatsheet https://github.com/rstudio/cheatsheets/blob/master/purrr.pdf
